---
title: "Lab: widgets for clustering"
author: Vince Carey
date: June 2, 2005
output: 
  ioslides_presentation:
    incremental: false
    fig_height: 4
runtime: shiny
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
suppressPackageStartupMessages({
suppressWarnings({
library(Biobase)
library(BiocStyle)
library(ggvis)
library(shiny)
library(rmarkdown)
library(cluster)
library(grid)
library(png)
library(limma)
library(fpc)
library(tissuesGeneExpression)
})
})
})
```
## How this lab works

- On the next page of this HTML document is an interface to cluster analysis of expression data on normal tissues 
- Controls are provided to vary 
    - the sample-to-sample distance used for hierarchical clustering
    - the agglomeration method used for building the tree
    - the number of clusters to be created by cutree
- On the page following the interface are a number of avenues by which the interface might be improved
- You can modify the .Rmd file to achieve these improvements


## normal tiss/gene exp. barcode (lab on next pg.)

```{r sta, echo=FALSE}
# this chunk sets up the expression data
data(tissuesGeneExpression)  # fragments brought to workspace
tiss = ExpressionSet(e)  # put them together
rownames(tab) = colnames(e)
pData(tiss) = tab
tiss = tiss[, tiss$SubType == "normal"]
#datatable(pData(tiss))
```

```{r lkwid, echo=FALSE}
# define a general interface for cluster analysis of columns of a matrix
hclustWidget = function(mat) {
 shinyApp(ui = fluidPage(  # lay out the interface
  fluidRow(
   column(3,  selectInput("distmeth", label = "Distance method:",
               choices = c("euclidean", "maximum", "manhattan",
               "binary"), selected = "euclidean")),

   column(3,  selectInput("fusemeth", label = "Agglomeration method:",
               choices = c("complete", "average", "ward.D2", "single",
                   "median", "centroid"), selected="complete")),
   column(3,  numericInput("numclus", label = "K:", 6, min = 2, max = 9)),
   column(3,  numericInput("ngenes", label = "N genes:", 500, min = 2, max = 100000))
          ),
  fluidRow(column(7, plotOutput("tree")), column(5, ggvisOutput("pcp")))
#
# define the data analysis steps -- this is a mess, there are probably
# redundancies
#
 ), server= function(input, output, session) {
#
# part 1: define the dendrogram and the Jaccard coefficients for
# a specific cut -- email the clusterboot guy and have him
# use message() instead of cat() and i can avoid the sink
#
    output$tree <- renderPlot({
dm = dist(mat[,1:input$ngenes], method=input$distmeth)
sink(tempfile())
cb <- clusterboot(dm, clustermethod=hclustCBI, method=input$fusemeth, k=input$numclus, showplots=FALSE, scaling=FALSE)
sink(NULL)
      dend = hclust( dm, method=input$fusemeth )
      plot(dend, main=paste0("Boot. Jacc. at k=", input$numclus, ": ",
        paste(round(cb$bootmean,2), collapse=", ")))
    })
#
# part 2: set up an ordination to PCs, and a tooltip on the ordination
#
    P1 <- reactive({
      pc = prcomp(mat[,1:input$ngenes])$x
      dm = dist(mat[,1:input$ngenes], method=input$distmeth)
      dend = hclust( dm, method=input$fusemeth )
      ct = cutree(dend, k=input$numclus)
      pcdf = data.frame(PC1=pc[,1], PC2=pc[,2], tiss=pData(tiss)$Tissue,
         rowid=1:nrow(pc), assigned=factor(ct))
#
# following for tooltip content
#
      all_values <- function(x) {
             if(is.null(x)) return(NULL)
             row <- pcdf[pcdf$rowid == x$rowid, ] # UGLY!
             paste0(names(row), ": ", format(row), collapse = "<br />")
           }
#
# ggvis is the only way i know to establish tooltips
#
      pcdf %>% ggvis(~PC1, ~PC2, key := ~rowid, fill = ~assigned) %>% layer_points() %>%
               add_tooltip(all_values, "hover") 
#      pairs(pc[,1:3], col=ct, pch=19, cex=1.5)
      }) 
      P1 %>% bind_shiny("pcp")
} )
}
#
# now build the matrix and invoke the interface
#
etiss = t(exprs(tiss))
rownames(etiss) = substr(tiss$Tissue,1,5)
hclustWidget(etiss)
```

## Exercise

- tisslab.Rmd can be processed with `library(shiny); run("tisslab.Rmd")`
- it should be improved in various ways
    - add links to the papers or GEO entries underlying the data
    - add to the available distance functions (include an approach to
setting the minkowski exponent, perhaps with a small number of fixed options);
also consider bioDist package
    - improve the content of the tooltip by getting rid of extraneous information and adding relevant information such as GEO identifier
    - replace the control labeled "N genes:" with a control that allows
selection of multiple named gene sets, perhaps defined by GO or KEGG or Reactome pathway terms 
    - suppose we find a set of gene sets that we like, how can we save the selection to the R session?
