---
title: "Monday Lab: Introduction to _R_ and _Bioconductor_"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteIndexEntry{Monday Lab: Introduction to R and Bioconductor}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Authors: Martin Morgan (<a
  href="mailto:mtmorgan@fhcrc.org">mtmorgan@fhcrc.org</a>), Sonali
  Arora (<a
  href="mailto:sarora@fredhutch.org">sarora@fredhutch.org</a>)<br />
Date: 15 June, 2015

This lab has four parts. Each part consists of informative material
followed by short or intermediate length exercises and case
studies. The material introduces or refreshes your knowledge of _R_,
and walks through some of the central classes and annotation resources
for working with high-throughput sequence data in _Bioconductor_. It's
recommended that you spend about 45 minutes on each of the following
sections. Take time to work through the code in the material as well
as the exercises, asking questions of your colleagues and the TAs as
needed.

- [Introduction to _R_](Lab-Monday-Introduction-R.html)
- [Introduction to _Bioconductor_](Lab-Monday-Introduction-Bioconductor.html)
- [Data Representations](Lab-Monday-Data-Representations.html)
- [Annotation](Lab-Monday-Annotation.html)
