---
title: "Lab: Performance and Parallel Evaluation"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteIndexEntry{Lab: Performance and Parallel Evaluation}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Authors: Martin Morgan (<a
  href="mailto:mtmorgan@fhcrc.org">mtmorgan@fhcrc.org</a>), Sonali
  Arora (<a
  href="mailto:sarora@fredhutch.org">sarora@fredhutch.org</a>)</a><br />
Date: 19 June, 2015

# Performance: deadly sins

The goal of this section is to learn to write correct, robust, simple
and efficient R code. We do this through a couple of case studies.

## Priorities

1. Correct: consistent with hand-worked examples (`identical()`, `all.equal()`)
2. Robust: supports realistic inputs, e.g., 0-length vectors, `NA`
   values, ...
3. Simple: easy to understand next month; easy to describe what it
   does to a colleague; easy to spot logical errors; easy to enhance.
4. Fast, or at least reasonable given the speed of modern computers.

## Strategies

1. Profile
   - _Look_ at the script to understand in general terms what it is doing.
   - _Step_ through the code to see how it is executed, and to gain an
     understanding of the speed of each line.
   - _Time_ evaluation of select lines or simple chunks of code with
     `system.time()` or the `r CRANpkg("microbenchmark")` package.
   - _Profile_ the code with a tool that indicates how much time is
     spent in each function call or line -- the built-in `Rprof()`
     function, or packages such as `r CRANpkg("lineprof")` or 
     `r CRANpkg("aprof")`
2. Vectorize -- operate on vectors, rather than explicit loops
    ```{r vectorize}
    x <- 1:10
    log(x)     ## NOT for (i in seq_along) x[i] <- log(x[i])
    ```
3. Pre-allocate memory, then fill in the result
    ```{r pre-allocate}
    result <- numeric(10)
    result[1] <- runif(1)
    for (i in 2:length(result))
           result[i] <- runif(1) * result[i - 1]
    result
    ```
4. Hoist common sub-expressions outside of repeated calculations, so
   that the sub-expression is only calculated once
   - Simple, e.g., 'hoist' constant multiplications from a `for` loop
   - Higher-level, e.g., use `lm.fit()` rather than repeatedly fitting
     the same design matrix.
5. Re-use existing, tested code
   - Efficient implementations of common operations -- `tabulate()`,
     `rowSums()` and friends, `%in%`, ...
   - Efficient domain-specific implementations, e.g., 
     `r Biocpkg("snpStats")` for GWAS linear models; `r Biocpkg("limma")`
     for microarray linear models; `r Biocpkg("edgeR")`, 
     `r Biocpkg("DESeq2")` for negative binomial GLMs relevant to
     RNASeq.

6. Re-think how to attack the problem
   - Different implementations
   - Alternative algorithms
7. Compile your script with the byte compiler
8. Use parallel evaluation
9. Speak in tongues -- 'foreign' languages like C, Fortran

## Case study: from iteration to vectorization

Here's an obviously inefficient function:
```{r inefficient}
f0 <- function(n, a=2) {
    ## stopifnot(is.integer(n) && (length(n) == 1) &&
    ##           !is.na(n) && (n > 0))
    result <- numeric()
    for (i in seq_len(n))
        result[[i]] <- a * log(i)
    result
}
```

Use `system.time()` to investigate how this algorithm scales with `n`,
focusing on elapsed time.

```{r system-time}
system.time(f0(10000))
n <- 1000 * seq(1, 20, 2)
t <- sapply(n, function(i) system.time(f0(i))[[3]])
plot(t ~ n, type="b")
```

Remember the current 'correct' value, and an approximate time

```{r correct-init}
n <- 10000
system.time(expected <- f0(n))
head(expected)
```

Revise the function to hoist the common multiplier, `a`, out of the
loop. Make sure the result of the 'optimization' and the original
calculation are the same. Use the `r CRANpkg("microbenchmark")`
package to compare the two versions

```{r hoist}
f1 <- function(n, a=2) {
    result <- numeric()
    for (i in seq_len(n))
        result[[i]] <- log(i)
    a * result
}
identical(expected, f1(n))

library(microbenchmark)
microbenchmark(f0(n), f1(n), times=5)
```

Adopt a 'pre-allocate and fill' strategy

```{r preallocate-and-fill}
f2 <- function(n, a=2) {
    result <- numeric(n)
    for (i in seq_len(n))
        result[[i]] <- log(i)
    a * result
}
identical(expected, f2(n))
microbenchmark(f0(n), f2(n), times=5)
```

Use an `*apply()` function to avoid having to explicitly pre-allocate,
and make opportunities for vectorization more apparent.

```{r use-apply}
f3 <- function(n, a=2)
    a * sapply(seq_len(n), log)

identical(expected, f3(n))
microbenchmark(f0(n), f2(n), f3(n), times=10)
```

Now that the code is presented in a single line, it is apparent that
it could be easily vectorized.  Seize the opportunity to vectorize it:

```{r use-vectorize}
f4 <- function(n, a=2)
    a * log(seq_len(n))
identical(expected, f4(n))
microbenchmark(f0(n), f3(n), f4(n), times=10)
```

Returning to our explicit iteration `f2()`, in these situations it
can be helpful to compile the code to a more efficient
representation. Do this using the compiler package.

```{r use-compiler}
library(compiler)
f2c <- cmpfun(f2)
n <- 10000
identical(f2(n), f2c(n))
microbenchmark(f2(n), f2c(n), times=10)
```

`f4()` definitely seems to be the winner. How does it scale with `n`?
(Repeat several times)

```{r vectorized-scale}
n <- 10 ^ (5:8)                         # 100x larger than f0
t <- sapply(n, function(i) system.time(f4(i))[[3]])
plot(t ~ n, log="xy", type="b")
```

Any explanations for the different pattern of response?

Lessons learned:

1. Vectorizing offers a huge improvement over iteration
2. Pre-allocate-and-fill is very helpful when explicit iteration is
   required.
3. `*apply()` functions help avoid need for explicit pre-allocation
   and make opportunities for vectorization more apparent. This may
   come at a small performance cost, but  is generally worth it
4. Hoisting common sub-expressions and using the _compiler_ package
   can be helpful for improving performance when explicit iteration is
   required.

## Case study: choosing algorithms

It can be very helpful to reason about an algorithm in an abstract
sense, to gain understanding about how an operation might
scale. Here's an interesting problem, taken from
[StackOverflow](http://stackoverflow.com/questions/16213029): Suppose
one has a very long **sorted** vector

```{r algo-init}
vec <- c(seq(-100,-1,length.out=1e6), rep(0,20), seq(1,100,length.out=1e6))
```

with the simple goal being to identify the number of values less than
zero. The original post and many responses suggested a variation of
scanning the vector for values less than zero, then summing

```{r algo-scan}
f0 <- function(v) sum(v < 0)
```

This algorithm compares each element of `vec` to zero, creating a
logical vector as long as the original, `length(v)`. This logical
vector is then scanned by `sum()` to count the number of elements
satisfying the condition.

Questions:

1. How many vectors of length `v` need to be allocated for this algorithm?
2. Based on the number of comparisons that need to be performed, how
   would you expect this algorithm to scale with the length of `v`?
   Verify this with a simple figure.
 
    ```{r algo-scan-time}
    N <- seq(1, 11, 2) * 1e6
    Time <- sapply(N, function(n) {
        v <- sort(rnorm(n))
        system.time(f0(v))[[3]]
    })
    plot(Time ~ N, type="b")
    ```

Is there a better algorithm, i.e., an approach that arrives at the
same answer but takes less time (and / or space)? The vector is
sorted, and we can take advantage of that by doing a _binary
search_. The algorithm is surprisingly simple: create an index to the
minimum (first) element, and the maximum (last) element. Check to
see if the element half way between is greater than or equal to
zero. If so, move the maximum index to that point. Otherwise, make
that point the new minimum. Repeat this procedure until the minimum
index is no longer less than the maximum index.

```{r algo-binary}
f3 <- function(x, threshold=0) {
    imin <- 1L
    imax <- length(x)
    while (imax >= imin) {
        imid <- as.integer(imin + (imax - imin) / 2)
        if (x[imid] >= threshold)
            imax <- imid - 1L
        else
            imin <- imid + 1L
    }
    imax
}
```

Approximately half of the possible values are discarded each
iteration, so we expect on average to arrive at the end after about
`log2(length(v))` iterations -- the algorithm scales with the log of
the length of `v`, rather than with the length of `v`, and no long
vectors are created. These difference become increasingly important as
the length of `v` becomes long.

Questions:

1. Verify with simple data that `f3()` and `f0()` result in
   `identical()` answers.
2. Compare how timing of `f3()` scales with vector length.
   
    ```{r algo-binary-perform}
    ## identity
    stopifnot(
        identical(f0((-2):2), f3((-2):2)),
        identical(f0(2:4), f3(2:4)),
        identical(f0(-(4:2)), f3(-(4:2))),
        identical(f0(vec), f3(vec)))
    
    ## scale
    N <- 10^(1:7)
    
    Time <- sapply(N, function(n) {
        v <- sort(rnorm(n))
        system.time(f3(v))[[3]]
    })
    plot(Time ~ N, type="b")
    ```

3. Use the `r CRANpkg("microbenchmark")` package to compare
   performance of `f0()` and `f3()` with the original data, `vec`.
4. R code can be compiled, and compilation helps most when doing
   non-vectorized operations like those in `f3()`. Use
   `compiler::cmpfun()` to compile `f3()`, and compare the result
   using microbenchmark.

    ```{r algo-relative}
    ## relative time
    library(microbenchmark)
    microbenchmark(f0(vec), f3(vec))
    
    library(compiler)
    f3c <- cmpfun(f3)
    microbenchmark(f3(vec), f3c(vec))
    ```

We could likely gain additional speed by writing the binary search
algorithm in C, but we are already so happy with the performance
improvement that we won't do that! 

It is useful to ask what is lost by `f3()` compared to `f0()`. For
instance, does the algorithm work on character vectors? What about
when the vector contains `NA` values? How are ties at 0 treated?

`findInterval()` is probably a better built-in way to solve the
original problem, and generalizes to additional situations. The idea
is to take the query that we are interested in, `0`, and find the
interval specified by `vec` in which it occurs.

```{r findInterval}
f4 <- function(v, query=0)
    findInterval(query - .Machine$double.eps, v)

identical(f0(vec), f4(vec))
microbenchmark(f0(vec), f3(vec), f4(vec))
```

The fact that it is flexible and well tested means that it would often
be preferred to `f3()`, even though it is less speedy. For instance,
compare the time it takes to query 10000 different points using `f3` and
iteration, versus `findInterval` and vectorization.

```{r findInterval-several}
threshold <- rnorm(10000)
identical(sapply(threshold, f3, x=vec), f4(vec, threshold))
microbenchmark(sapply(x, f3), f4(vec, x))
```

Some R functions that implement efficient algorithms are `sort()`
(including radix sort), `match()` (hash table look-up), and
`tabulate()`; these can be useful in your own code. 

Lessons learned:

1. Choice of algorithm can be very important
2. Implementing classical algorithms (like binary search) can be a
   rewarding learning experience even when, at the end of the day, it
   may be better to use existing functions.
3. The built-in R functions that implement efficient algorithms can be
   important building-blocks for more complicated code.

# Parallel evaluation

## Types of parallel problems

### 'Embarassingly' parallel

### Other

Sequential, e.g., input + transform

- An input 'reader' dedicated to sequentially inputting chunks of data
- Several 'workers' responsible for transforming each chunk
- Appropriate when work done is expensive relative to time required
  for input

Multiple program, single data

- Each data chunk processed by several different 'programs'. Naive
  example: calculating mean and median could represent two separate
  'programs'
- Unusual in data analysis problems

## Scale of computation

Cores

- _Forked_ processes
- Independent processes

Clusters

Clouds

## _GenomicFiles_

## _BiocParallel_

Embarassingly parallel iteration or vectorization

```{r bplapply-bpvec-FIXME}
```

Registering alternative 'back ends'

### Coming soon (available in 'devel')

Error handing

Logging

### Why not some other solution

Other solutions

- _parallel_ (base R)
- `r CRANpkg("foreach")`
- `r CRANpkg("BatchJobs")`

_foreach_

- A nice and intuitive interface, especially for non-R programmers.
- Nice plug-in model, so easy to adapt code to different computing
  environments, e.g., forked cores or independent processes
  
    ```{r foreach-fork-FIXME}
    ```

- The `foreach()` paradigm emphasizes iteration, which as we've seen
  requires that the user avoid poor programming styles (e.g.,
  pre-allocate and fill, rather than copy-and-append) and can be an
  obstacle to vectorization

_BatchJobs_

- Probably a good solution for developing large-scale projects that
  require use of clusters.
- Each cluster is different, so very difficult to write code that is
  portable: difficult to share your analysis with others.
