---
title: "Monday Lab, part A: Introduction to _R_"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteIndexEntry{Monday Lab, part A: Introduction to R}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Authors: Martin Morgan (<a
  href="mailto:mtmorgan@fhcrc.org">mtmorgan@fhcrc.org</a>), Sonali
  Arora (<a
  href="mailto:sarora@fredhutch.org">sarora@fredhutch.org</a><br />
Date: 15 June, 2015<br />
Back to [Monday labs](Lab-Monday-Introduction.html)

# R

Language and environment for statistical computing and graphics

- Full-featured programming language
- Interactive and *interpretted* -- convenient and forgiving
- Coherent, extensive documentation
- Statistical, e.g. `factor()`, `NA`
- Extensible -- CRAN, Bioconductor, github, ...

Vector, class, object

- Efficient _vectorized_ calculations on 'atomic' vectors `logical`,
  `integer`, `numeric`, `complex`, `character`, `byte`
- Atomic vectors are building blocks for more complicated _objects_
  - `matrix` -- atomic vector with 'dim' attribute
  - `data.frame` -- list of equal length atomic vectors
- Formal _classes_ represent complicated combinations of vectors,
  e.g., the return value of `lm()`, below

Function, generic, method

- Functions transform inputs to outputs, perhaps with side effects,
  e.g., `rnorm(1000)`
  - Argument matching first by name, then by position
  - Functions may define (some) arguments to have default values
- _Generic_ functions dispatch to specific _methods_ based on class of
  argument(s), e.g., `print()`. 
- Methods are functions that implement specific generics, e.g.,
  `print.factor`; methods are invoked _indirectly_, via the generic.
- Many but not all functions able to manipulate a particular class are
  methods, e.g., `abline()` used below is a plain-old-funciton.

Introspection

- General properties, e.g., `class()`, `str()`
- Class-specific properties, e.g., `dim()`

Help

- `?print`: help on the generic print 
- `?print.data.frame`: help on print method for objects of class
    data.frame.

Example

```{r}
x <- rnorm(1000)                   # atomic vectors
y <- x + rnorm(1000, sd=.5)
df <- data.frame(x=x, y=y)         # object of class 'data.frame'
plot(y ~ x, df)                    # generic plot, method plot.formula
fit <- lm(y ~x, df)                # object of class 'lm'
methods(class=class(fit))          # introspection
anova(fit)
plot(y ~ x, df)
abline(fit, col="red", lwd=3, lty=2) # a function, not generic.method
```


# Case study: ALL phenotypic data

These case studies servers as refreshers on _R_ input and manipulation
of data.

Input a file that contains ALL (acute lymphoblastic leukemia) patient
information

```{r echo=TRUE, eval=FALSE}
fname <- file.choose()   ## "ALLphenoData.tsv"
stopifnot(file.exists(fname))
pdata <- read.delim(fname)
```
```{r echo=FALSE}
fname <- system.file(package="CSAMA15Morgan", "extdata",
    "ALLphenoData.tsv")
stopifnot(file.exists(fname))
pdata <- read.delim(fname)
```

Check out the help page `?read.delim` for input options, and explore
basic properties of the object you've created, for instance...

```{r ALL-properties}
class(pdata)
colnames(pdata)
dim(pdata)
head(pdata)
summary(pdata$sex)
summary(pdata$cyto.normal)
```

Remind yourselves about various ways to subset and access columns of a
data.frame

```{r ALL-subset}
pdata[1:5, 3:4]
pdata[1:5, ]
head(pdata[, 3:5])
tail(pdata[, 3:5], 3)
head(pdata$age)
head(pdata$sex)
head(pdata[pdata$age > 21,])
```

It seems from below that there are 17 females over 40 in the data set,
but when sub-setting `pdata` to contain just those individuals 19 rows
are selected. Why? What can we do to correct this?

```{r ALL-subset-NA}
idx <- pdata$sex == "F" & pdata$age > 40
table(idx)
dim(pdata[idx,])
```

Use the `mol.biol` column to subset the data to contain just
individuals with 'BCR/ABL' or 'NEG', e.g.,

```{r ALL-BCR/ABL-subset}
bcrabl <- pdata[pdata$mol.biol %in% c("BCR/ABL", "NEG"),]
```

The `mol.biol` column is a factor, and retains all levels even after
subsetting. How might you drop the unused factor levels?

```{r ALL-BCR/ABL-drop-unused}
bcrabl$mol.biol <- factor(bcrabl$mol.biol)
```

The `BT` column is a factor describing B- and T-cell subtypes

```{r ALL-BT}
levels(bcrabl$BT)
```

How might one collapse B1, B2, ... to a single type B, and likewise for T1, T2, ..., so there are only two subtypes, B and T

```{r ALL-BT-recode}
table(bcrabl$BT)
levels(bcrabl$BT) <- substring(levels(bcrabl$BT), 1, 1)
table(bcrabl$BT)
```

Use `xtabs()` (cross-tabulation) to count the number of samples with
B- and T-cell types in each of the BCR/ABL and NEG groups

```{r ALL-BCR/ABL-BT}
xtabs(~ BT + mol.biol, bcrabl)
```

Use `aggregate()` to calculate the average age of males and females in
the BCR/ABL and NEG treatment groups.

```{r ALL-aggregate}
aggregate(age ~ mol.biol + sex, bcrabl, mean)
```

Use `t.test()` to compare the age of individuals in the BCR/ABL versus
NEG groups; visualize the results using `boxplot()`. In both cases,
use the `formula` interface. Consult the help page `?t.test` and re-do
the test assuming that variance of ages in the two groups is
identical. What parts of the test output change?

```{r ALL-age}
t.test(age ~ mol.biol, bcrabl)
boxplot(age ~ mol.biol, bcrabl)
```

# Case study: weighty matters

This case study is a second walk through basic data manipulation and
visualization skills.  We use data from the US Center for Disease
Control's Behavioral Risk Factor Surveillance System ([BRFSS][])
annual survey. Check out the web page for a little more
information. We are using a small subset of this data, including a
random sample of 10000 observations from each of 1990 and 2010.

Input the data using `read.csv()`, creating a variable `brfss` to hold
it.  Use `file.choose()` to locate the data file BRFSS-subset.csv

```{r echo=TRUE, eval=FALSE}
fname <- file.choose()   ## BRFSS-subset.csv
stopifnot(file.exists(fname))
brfss <- read.csv(fname)
```
```{r echo=FALSE}
fname <- system.file(package="CSAMA15Morgan", "extdata",
    "BRFSS-subset.csv")
stopifnot(file.exists(fname))
brfss <- read.csv(fname)
```

## Exercises: base plotting functions

1. Explore the data using `class()`, `dim()`, `head()`, `summary()`,
   etc. Use `xtabs()` to summarize the number of males and females in
   the study, in each of the two years.

2. Use `aggregate()` to summarize the average weight in each sex and
   year.
  
3. Create a scatterplot showing the relationship between the square
   root of weight and height, using the `plot()` function and the
   `main` argument to annotate the plot. Note the transformed
   Y-axis. Experiment with different plotting symbols (try the command
   `example(points)` to view different points).

   ```{r brfss-simple-plot}
   plot(sqrt(Weight) ~ Height, brfss, main="All Years, Both Sexes")
   ```
   
4. Color the female and male points differently. To do this, use the
   `col` argument to `plot()`. Provide as a value to that argument a
   vector of colors, subset by `brfss$Sex`.

5. Create a subset of the data containing only observations from
   2010.

   ```{r brfss-subset}
   brfss2010 <- brfss[brfss$Year == "2010", ]
   ```

6. Create the figure below (two panels in a single figure). Do this by
   using the `par()` function with the `mfcol` argument before calling
   `plot()`. You'll need to create two more subsets of data, perhaps
   when you are providing the data to the function `plot`.

   ```{r brfss-pair-plot}
   opar <- par(mfcol=c(1, 2))
   plot(sqrt(Weight) ~ Height, brfss2010[brfss2010$Sex == "Female", ],
        main="2010, Female")
   plot(sqrt(Weight) ~ Height, brfss2010[brfss2010$Sex == "Male", ],
        main="2010, Male")
   par(mfcol=c(1, 1))
   ```

7. Plotting large numbers of points means that they are often
   over-plotted, potentially obscuring important patterns. Experiment
   with arguments to `plot()` to address over-plotting, e.g.,
   `pch='.'` or `alpha=.4`. Try using the `smoothScatter()` function
   (the data have to be presented as `x` and `y`, rather than as a
   formula). Try adding the [hexbin][] library to your R session
   (using `library()`) and creating a `hexbinplot()`.

## Exercises: Trellis graphics and the _lattice_ package

R has a number of additional plotting facilities, both as part of the
'base' distribution and user-contributed packages. The _lattice_
package adopts a particular philosophy about the presentation of data,
and can be a very effective visualization tool.

1.  Use `library()` to load the _lattice_ package.

    ```{r lattice}
    library(lattice)
    ```

2. Create the following figure using the `xyplot()` function with a
   formula and the `brfss2010` data. The formula is `sqrt(Weight) ~
   Height | Sex`, which can be read as `square root of Weight as a
   function of Height, conditioned on Sex'.

   ```{r lattice-conditioning}
   xyplot(sqrt(Weight) ~ Height | Sex, brfss2010)
   ```

3. Add a background grid and a regression line to each panel using the
   argument `type=c('p', 'g', 'r')`; change the width (`lwd`) and
   color (`col.line`) of the regression line. For some hints on other
   arguments, see the help pages `?xyplot` (for overall structure of
   the plot) and `?panel.xyplot` (for how each 'panel' of data is
   plotted).
  
4. Create the following figure. Use the `densityplot()` function with
   the formula `~ sqrt(Weight)`. The `group=Sex` function argument
   creates separate lines for each sex. Be sure to use
   `plot.points=FALSE` to avoid a `rug' of points at the base of the
   figure. Can you add a key (legend)?
   
   ```{r lattice-density}
   densityplot(~sqrt(Weight), brfss2010, group=Sex, plot.points=FALSE)
   ```

5. Create the figure below using the `bwplot` function. The formula
   requires that `Year` be coerced to a `factor`, `factor(Year)`.

    ```{r lattice-bwplot}
    bwplot(sqrt(Weight) ~ factor(Year) | Sex, brfss)
    ```
  
6.  Create the Figure below, a _violin_ plot, using `bwplot()` and the
    `panel` argument set to `panel.violin`. from `?bwplot` we learn
    that `panel` is a function that determines how each panel is
    drawn; the details for controling the violin panel plot are
    described on the help page `?panel.violin`.

    ```{r lattice-violin}
    bwplot(sqrt(Weight) ~ factor(Year) | Sex, brfss, panel=panel.violin)
    ```

7. (Advanced) We can write our own `panel` argument to _lattice_
   functions to influence how each panel is displayed. Here we add a
   point at the median age and weight.

    ```{r lattice-panel}
    xyplot(sqrt(Weight) ~ Height|Sex, brfss2010,
        panel = function(x, y, ...) {
            panel.xyplot(x, y, ...)
            panel.points(median(x, na.rm=TRUE), median(y, na.rm=TRUE), 
                cex=2, pch=20, col="red")
        },
        type=c("p", "g", "r"), lwd=2, col.line="red", xlim=c(120, 210))
    ```

[BRFSS]: http://www.cdc.gov/brfss/

[biocViews]: http://bioconductor.org/packages/BiocViews.html#___Software
[AnnotationData]: http://bioconductor.org/packages/BiocViews.html#___AnnotationData

[aprof]: http://cran.r-project.org/web/packages/aprof/index.html
[hexbin]: http://cran.r-project.org/web/packages/hexbin/index.html
[lineprof]: https://github.com/hadley/lineprof
[microbenchmark]: http://cran.r-project.org/web/packages/microbenchmark/index.html


[AnnotationDbi]: http://bioconductor.org/packages/AnnotationDbi.html
[BSgenome]: http://bioconductor.org/packages/BSgenome.html
[Biostrings]: http://bioconductor.org/packages/Biostrings.html
[CNTools]: http://bioconductor.org/packages/CNTools.html
[ChIPQC]: http://bioconductor.org/packages/ChIPQC.html
[ChIPpeakAnno]: http://bioconductor.org/packages/ChIPpeakAnno.html
[DESeq2]: http://bioconductor.org/packages/DESeq2.html
[DiffBind]: http://bioconductor.org/packages/DiffBind.html
[GenomicAlignments]: http://bioconductor.org/packages/GenomicAlignments.html
[GenomicRanges]: http://bioconductor.org/packages/GenomicRanges.html
[IRanges]: http://bioconductor.org/packages/IRanges.html
[KEGGREST]: http://bioconductor.org/packages/KEGGREST.html
[PSICQUIC]: http://bioconductor.org/packages/PSICQUIC.html
[Rsamtools]: http://bioconductor.org/packages/Rsamtools.html
[ShortRead]: http://bioconductor.org/packages/ShortRead.html
[VariantAnnotation]: http://bioconductor.org/packages/VariantAnnotation.html
[VariantFiltering]: http://bioconductor.org/packages/VariantFiltering.html
[VariantTools]: http://bioconductor.org/packages/VariantTools.html
[biomaRt]: http://bioconductor.org/packages/biomaRt.html
[cn.mops]: http://bioconductor.org/packages/cn.mops.html
[h5vc]: http://bioconductor.org/packages/h5vc.html
[edgeR]: http://bioconductor.org/packages/edgeR.html
[ensemblVEP]: http://bioconductor.org/packages/ensemblVEP.html
[limma]: http://bioconductor.org/packages/limma.html
[metagenomeSeq]: http://bioconductor.org/packages/metagenomeSeq.html
[phyloseq]: http://bioconductor.org/packages/phyloseq.html
[snpStats]: http://bioconductor.org/packages/snpStats.html

[org.Hs.eg.db]: http://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html
[TxDb.Hsapiens.UCSC.hg19.knownGene]: http://bioconductor.org/packages/release/data/annotation/html/TxDb.Hsapiens.UCSC.hg19.knownGene.html
[BSgenome.Hsapiens.UCSC.hg19]: http://bioconductor.org/packages/release/data/annotation/html/BSgenome.Hsapiens.UCSC.hg19.html
