---
title: "Interactive data visualization using Shiny"
author: "Andrzej Oleś"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output: ioslides_presentation
runtime: shiny
---

## What is Shiny?

Web application framework for R from RStudio

<p align="center">http://shiny.rstudio.com/</p>

Shiny combines the computational power of R with the interactivity of the modern web

- build interactive web applications with R easily
- no HTML or JavaScript knowledge required
- reactive programming model:<br/>
  output changes instantly on input modification

## Slide with Interactive Plot

```{r, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Code and Output

```{r}
summary(cars)
```


