---
title: "Clustering and classification"
author: Vince Carey
date: May 22, 2005
output: 
  ioslides_presentation:
    incremental: true
    fig_height: 4
runtime: shiny
---

```{r setup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
suppressWarnings({
library(shiny)
library(rmarkdown)
library(yeastCC)
library(cluster)
library(grid)
library(png)
library(limma)
library(rafalib)
})
})
```

## Basic concepts {.build}

- Organisms are assayed on multiple features
- _Variability_ in feature measures exhibits _structure_
- Clustering:
    - For some grouping, between-group variation is larger than within-group variation
    - Our goals are to find, evaluate, and interpret such groupings 
- Classification:
    - Organisms are sorted into classes and labeled
    - Rules for classification are maps from features to labels
    - Our goals are to find, evaluate, and interpret such rules

## Two basic approaches to clustering, agglomerative and divisive {.flexbox .vcenter}
```{r dofigagg,fig=TRUE,echo=FALSE,fig.height=3.9}
par(mfrow=c(1,2))
pp = pam(dist(iris[,1:4]), k=3)
mhc = function(...) {ans = hclust(...); ans$labels=substr(iris[,5],1,1); ans}
myplclust(mhc(dist(iris[,1:4]), method="average"), main="hclust/euc/avg on iris", lab.col=as.numeric(factor(iris[,5])))
plot(pp, which=1, main="pam on iris, k=3")
```

- Fisher's iris data: three labeled species, four anatomic features

## Genomic example: gene expression in the cell cycle of budding yeast

```{r dofigyy,fig=TRUE,echo=FALSE,fig.height=5.5}
library(png)
im = readPNG("fullcyc.png")
grid.raster(im)
```


## a data extract: _S. cerevisiae_ colony synchronized with alpha pheromone

```{r lky}
library(yeastCC)
data(spYCCES)
alp = spYCCES[, spYCCES$syncmeth=="alpha"]
rbind(time=alp$time[1:5],exprs(alp)[1:5,1:5])
```

- G=`r nrow(exprs(alp))` genes comprise rows, N=`r ncol(exprs(alp))` timed samples comprise columns

## A pattern of interest

- Define a _basal oscillator_ to be a gene with expression varying with the cell cycle in a specific way
    - for alpha-synchronized colony, cell cycle period is about 66 minutes
    
- Theoretical expression trajectory
```{r lkba,fig=TRUE,echo=FALSE}
times=0:132
bo = sin(2*pi*times/66)
plot(times, bo, type="l")
abline(h=0,lty=2,col="gray")
abline(v=66,lty=2,col="gray")
```

## Can you give a mathematical definition of a basal oscillator?

## Formalism
 
- $t$ denotes time (in minutes) elapsed from synchronization
- $X_g(t)$ denotes reported expression of gene $g$ at time $t$
- $m_g = \min_t X_g(t)$, $M_g = \max_t X_g(t)$
- $U_g(t) = 2 \times [\frac{X_g(t) - m_g}{M_g-m_g} - 0.5]$ is
signed fractional excursion (sfe) of gene $g$ at time $t$
    - if $g$ is at minimal reported value at $t$, $U_g(t) = -1$
    - if $g$ is at maximal reported value at $t$, $U_g(t) = +1$
- if $g$ is a basal oscillator, what is the form of $U_g(t)$?

## One possible form for $U_g(t)$ for $g$ a basal oscillator

- $U_g(t) = \sin(2 \pi t/ 66)$, $t$ in minutes from synchronization
- Why?
    - $\sin$ function has range [-1,1]
    - smoothness corresponds to gradual nature of transition
    - returns to 0 at multiples of 66 minutes
- Drawbacks?
    - periodicity is biologically motivated, but the detailed trajectory with various local symmetries is not justified 
- Survey: How many genes are reasonably modeled by the basal oscillator pattern? 0, 10, 100, 1000?

## Measuring similarity of observed expression trajectories to the basal oscillator pattern

- High-school analytic geometry: distance between two points in $R^3$
- $p_1 = (x_1, y_1, z_1)$, $p_2 = (x_2, y_2, z_2)$
- $\Delta x = x_1 - x_2$, etc.
- $d(p_1, p_2) = \sqrt{(\Delta x)^2 + (\Delta y)^2 + (\Delta z)^2}$
- "Euclidean distance"

## Application of the distance concept

- What is the dimension of the space in which a yeast gene expression trajectory resides?
- How can we define the distance between a given gene's trajectory and that corresponding to the basal oscillator pattern?  Assumptions?
-
```{r f2,fig=TRUE,echo=FALSE}
suppressWarnings({
times=1:132
bo = sin(2*pi*times/66)
ea = exprs(alp)
unitize = function(x) {
 tmp = (x-min(x,na.rm=TRUE))/(max(x,na.rm=TRUE)-min(x,na.rm=TRUE))
 2*(tmp-.5)
}
uea = t(apply(ea,1,unitize))
plot(times, bo, type="l", xlab="time", ylab="sfe")
lines(alp$time, uea[1,], lty=2)
legend(38,.87, lty=c(1,2), legend=c("basal osc.", featureNames(alp)[1]))
})
```

## Computing distances to basal oscillator pattern

```{r dodis}
bot = function(tim) sin(2*pi*tim/66)
bo = bot(alp$time)
d2bo = function(x) sqrt(sum((x-bo)^2))
suppressWarnings({ds = apply(uea, 1, d2bo)})
md = which.min(ds)
md
summary(ds)
```

## The nearest gene

```{r lkgcl, fig=TRUE}
plot(alp$time, bo, type="l", xlab="time", ylab="sfe")
lines(alp$time, uea[md,], lty=2)
legend(38,.87, lty=c(1,2), legend=c("basal osc.", featureNames(alp)[md]))
```

## The distribution of distances

```{r lkh,fig=TRUE}
hist(ds, xlab="Euclidean distance to basal oscillator pattern")
```

- What should this look like if there is clustering?

## "Top ten!"

```{r lkgcl2, fig=TRUE}
todo = names(sort(ds))[1:10]
plot(alp$time, bo, type="l", xlab="time", ylab="sfe")
for (md in todo) lines(alp$time, uea[md,], lty=2, col="gray")
#legend(38,.87, lty=c(1,2), legend=c("basal osc.", featureNames(alp)[md]))
```

## Is it a cluster?

- Recall the definition:
    - _Variability_ in feature measures exhibits _structure_
    - For some grouping, between-group variation is larger than within-group variation
    - Can we make this more precise?

## Definition from ?silhouette
```
   For each observation i, the _silhouette width_ s(i) is defined as
     follows:
     Put a(i) = average dissimilarity between i and all other points of
     the cluster to which i belongs (if i is the _only_ observation in
     its cluster, s(i) := 0 without further calculations).  For all
     _other_ clusters C, put d(i,C) = average dissimilarity of i to all
     observations of C.  The smallest of these d(i,C) is b(i) := \min_C
     d(i,C), and can be seen as the dissimilarity between i and its
     “neighbor” cluster, i.e., the nearest one to which it does _not_
     belong.  Finally,

                   s(i) := ( b(i) - a(i) ) / max( a(i), b(i) ).         
     
     ‘silhouette.default()’ is now based on C code donated by Romain
     Francois (the R version being still available as
     ‘cluster:::silhouette.default.R’).
```


## Realizations of an unstructured grouping scheme
- Form some arbitrarily chosen groups of size ten
- The code:
```{r dospl}
allf = featureNames(alp)
set.seed(2345)
scramble = function(x) sample(x, size=length(x), replace=FALSE) 
cands = scramble(setdiff(allf, todo))[1:40]
sc = split(cands, gids <- rep(2:5,each=10))
ml = lapply(sc, function(x) uea[x,])
```

## Trajectories from the arbitrary groups {.flexbox .vcenter}

```{r getgr, fig=TRUE,echo=FALSE,fig.height=5}
par(mfrow=c(2,2), mar=c(4,2,2,2))
for (i in 1:4) {
 plot(alp$time, ml[[i]][1,], type="l", lty=2, col="gray", main=paste("group", i+1), xlab="time", ylab="sfe")
 for (j in 2:9) lines(alp$time, ml[[i]][j,], type="l", lty=2, col="gray")
}
```

## The silhouette plot

```{r dof,fig=TRUE,echo=FALSE}
alld = c(todo, cands)
labs = c(rep(1,10), gids)
d = dist(uea[alld,])
s = silhouette(labs, d) 
plot(s)
```

## Recap
- A target pattern was defined, using knowledge of the cell cycle period
- Expression patterns were transformed to conform to the dynamic
range of the target pattern
- A distance function was defined and genes ordered by proximity to
target
- A group of ten genes nearest to target trajectory was identified
and compared to arbitrarily formed groups using silhouette widths
- Transformation, Distance, and Comparison are fundamental elements
of all cluster analysis
    - but a fixed target pattern is not so common in genomics
    - compare handwriting or speech waveform analysis

## Quiz
- Give the mathematical definition of the hyperbasal oscillator pattern,
that has value 1 at time 0 and returns to 1 at 66 minutes
- Which gene has expression trajectory closest to that of the
hyperbasal oscillator?
- How many steps to determine the mean silhouette width for the
ten genes closest to the hyperbasal pattern?

## Solution

```{r dodis2,fig=FALSE}
hbot = function(tim) cos(2*pi*tim/66)
hbo = hbot(alp$time)
d2hbo = function(x) sqrt(sum((x-hbo)^2))
suppressWarnings({cds = apply(uea, 1, d2hbo)})
cmd = which.min(cds)
cmd
summary(cds)
```

## The most hyperbasal gene

```{r lkgcl3, fig=TRUE}
plot(alp$time, hbo, type="l", xlab="time", ylab="sfe")
lines(alp$time, uea[cmd,], lty=2)
legend(20,.87, lty=c(1,2), legend=c("basal osc.", featureNames(alp)[cmd]))
```

## "Top ten!"

```{r getmoc,fig=TRUE}
ctodo = names(sort(cds)[1:10])
plot(alp$time, hbo, type="l", xlab="time", ylab="sfe")
for (md in ctodo) lines(alp$time, uea[md,], lty=2, col="gray")
```

## Silhouette continuation

```{r cdof,fig=TRUE,echo=FALSE}
alld = c(todo, cands, ctodo)
labs = c(rep(1,10), gids, rep(6,10))
d = dist(uea[alld,])
s = silhouette(labs, d) 
plot(s)
```

## Caveats

- We ignored the natural units reported for expression
- The pure sinusoid need not be a reasonable trajectory model for any gene
- We arbitrarily thresholded to achieve groups of size 10
- Can we use the data to define groups of genes with similar expression patterns without so much conceptual intervention?

## Hierarchical clustering

- Subdivision of the genome into coexpressed groups is of general interest
- Agglomerative algorithms use the distance measure to combine very similar genes into new entities 
- The process repeats until only one entity remains

## Filtering

- We would like to look for structure among expression patterns of genes exhibiting oscillatory behavior
- Thus we would like to filter away genes with non-periodic trajectories
- Trigonometric regression can help
    - $t$ is transformed to [0,1]; estimate $s_{gj}$ and $c_{gj}$ in
$$ X_g(t) = \sum_{j=1}^J s_{gj} \sin 2j\pi t + \sum_{j=1}^J c_{gj} \cos 2j \pi t + e_g(t) $$
- We will cluster genes possessing relatively large $F$ statistics for this model, fixing $J=2$

## limma for trigonometric regression fits

```{r dolimm}
options(digits=2)
x = (alp$time %% 66)/66
mm = model.matrix(~sin(2*pi*x)+cos(2*pi*x)+sin(4*pi*x)+cos(4*pi*x)-1)
colnames(mm) = c("s1", "c1", "s2", "c2")
library(limma)
m1 = lmFit(alp, mm)
em1 = eBayes(m1)
topTable(em1, 1:4, n=5)
```

## Interactive interface


```{r tryint,echo=FALSE}
top100 = topTable(em1, 1:4, n=100)
format_num <- function(col) {
  if (is.numeric(col))
    sprintf('%1.3f', col)
  else
    col
}
top100f = as.data.frame(lapply(top100, format_num), stringsAsFactors=FALSE)
top100f = cbind(orf=rownames(top100),top100f)
rownames(top100f) = rownames(top100)
top100f[,2] = as.numeric(top100f[,2])
top100f[,3] = as.numeric(top100f[,3])
top100f[,4] = as.numeric(top100f[,4])
top100f[,5] = as.numeric(top100f[,5])

ui = fluidPage(titlePanel("trigonometric regression for yeast expression"),
#
# should be selectize
#
   fluidRow( column(3, DT::dataTableOutput('toptable')) ,
   column(3,selectizeInput('orf', 'ORF', choices=NULL,
           selected="YMR011W", options=list(placeholder="YMR011W"),
           multiple=FALSE)),
   column(5, plotOutput('trajplot'))
   )
  )

server = function(input, output, session) {

# Apply the function to each column, and convert the list output back to a data frame
   INI = reactive({
      if (is.null(input$orf) | nchar(input$orf)==0) return("YMR011W")
      else input$orf
      })
   output$toptable = DT::renderDataTable(
      DT::datatable(top100f[,c(1,2,3,4,5,8)])
      )
   output$trajplot = renderPlot( {
      plot(alp$time, exprs(alp)[INI(),], main=INI(), xlab="time", ylab="reported expr.")
      co = as.numeric(top100f[INI(),2:5])
      ti = ((0:132)%%66)/66
      pred = co[1]*sin(2*pi*ti)+co[2]*cos(2*pi*ti)+co[3]*sin(4*pi*ti)+
            co[4]*cos(4*pi*ti)
      lines(0:132, pred, lty=2) }
     )
   updateSelectizeInput(session, 'orf', choices = rownames(top100), server = TRUE)
}

shinyApp(ui=ui, server=server)
```
