---
title: "Plotting Regions from BAM files directly"
author: "Paul Theodor Pyl"
date: "18 Jun 2015"
output: html_document
---

## Setting Up

First we will need to get some R packages from the local repository (At home we can just use `biocLite` to install those).

```{r, eval = FALSE}
source("http://192.168.0.9/setRepos.R") 
install.packages(c("h5vc", "h5vcData"))
install.packages("BSgenome.Hsapiens.NCBI.GRCh38") # We need this genome to define our reference, you should use a genome that corresponds to the bam files you want to use
```

Typically an HDF5 tally file will store the nucleotide tally data and the accompanying sample data and the functions for plotting defined in the `h5vc` package are. If we do not want to create an HDF5 file representation of our data, we can define a new function that makes sure that the data and sample data are present in the correct format.

First we load the required packages, `h5vc` for plotting and `BSgenome.Hsapiens.NCBI.GRCh38` to provide the reference genome sequence. We use `BSgenome.Hsapiens.NCBI.GRCh38` because the example `.bam` files that I will use here are aligned against that revision of the human reference genome. You should use an appropriate genome object here, e.g. if you want to work in *Arabidopsis Thaliana* instead you would maybe use the `BSgenome.Athaliana.TAIR.TAIR9` package instead.

```{r, echo=TRUE, message=FALSE}
require(h5vc)
require(BSgenome.Hsapiens.NCBI.GRCh38)
```

Next we find our example data `.bam` files and create a `sampleData` object that is simply a `data.frame` describing the samples we want to work with.

```{r}
files <- list.files( system.file("extdata", package = "h5vcData"), "Pt.*bam$" )
bamFiles <- file.path( system.file("extdata", package = "h5vcData"), files)

sampleData <- data.frame(
  SampleFiles = files,
  Sample = sapply(strsplit(files, split = "\\."), function(x) x[1]),
  Type = rep(c("Case", "Control"), length(files)/2),
  Column = seq(length(files)),
  Patient = substr(files, 1, 4),
  stringsAsFactors = FALSE #Silly, but important to get the track labels right
  )

sampleData
```

## Defining the plotting function

Finally we define the `plotBAMs` function that will take a `GRanges` object describing the genomic intervals we want to plot, the list of `.bam` files, the `sampleData` object and the reference object (that will be `BSgenome.Hsapiens.NCBI.GRCh38` in our case).

```{r}
plotBAMs <- function( ranges, bamFiles, sampleData, reference){
  require(h5vc)
  theData <- tallyRanges( bamFiles, ranges = ranges, reference = reference )
  for( idx in seq(length(ranges))){ # This is needed since we dont use HDF5 to store the data
    theData[[idx]]$h5dapplyInfo <- list( Blockstart = start(ranges[idx]), Blockend = end(ranges[idx]))
  }
  mismatchPlot(
    theData, sampleData
    )
}
```

## Creating plots of example data for the DNMT3A gene

Here we define the genomic ranges we will want to plot, they correspond to exons of the DNMT3A gene in our example. The `GRanges` should come from a VCF file with variant calls when we want to visualise variant calls, but for the sake of this example the DNMT3A exons will suffice.

```{r}
require(GenomicRanges)
dnmt3a <- read.table(system.file("extdata", "dnmt3a.txt", package = "h5vcData"), header=TRUE, stringsAsFactors = FALSE)
dnmt3a <- with( dnmt3a, GRanges(seqname, ranges = IRanges(start = start, end = end)))
dnmt3a <- reduce(dnmt3a)
dnmt3a
```

Let's plot some DNMT3A exons  (2,3 and 4) in the first 4 samples (the object `Hsapiens` is defined in the `BSgenome.Hsapiens.NCBI.GRCh38` package and contains the genomic reference sequence2):

```{r}
p <- plotBAMs(
  ranges = dnmt3a[2:4],
  bamFiles = bamFiles[1:4],
  sampleData = sampleData[1:4,],
  reference = Hsapiens
  )
print(p)
```

`p` is a ggplot object, so we can modify how it looks by adding theme commands and so on:
```{r}
print(p + theme(text = element_text(colour = "hotpink"))) 
```

Hmm, maybe making the text pink wasn't the best idea, let's try something else:

```{r, fig.height=10, fig.width=16}
print(p + theme(text = element_text(size = 16), axis.text.x = element_text(angle = 45, size = 12, vjust = 1, hjust = 1), panel.margin.x = unit(1, units = "cm"))) 
```

## Exercises

1. Apply your knowledge of creating shiny apps to create a browser-app. This app should let you select which samples and exons (of DNMT3A) should be plotted and then display the selected region.
2. Use your knowledge of the `VariantAnnotation` package to implement a shiny app or set of functions that loads a VCF file and plots ranges around the variants. You might need a set of appropriate `.bam` files to do this, i.e. `.bam` files that contain the data that gave rise to the variant calls in your VCF file.

